---
title: "raincloud plot, modified with reporting S.E. of the group"
output:
  pdf_document: default
  html_document: default
---
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.3368186.svg)](https://doi.org/10.5281/zenodo.3368186)

### A fun attempt to play with Raincloud Plots!

First, import the data here
```{r setup, include=FALSE, cache=TRUE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

##code to read the csv data
file1 <- read.csv("../motivation_gamble_data.csv", header=TRUE, sep=",")

##select only data from 'study1' as an example here
df <- subset(file1, expt=="study1")

#to have a glimpse of the data
head(df,5)

```



Set up the needed libraries
```{r, include=FALSE}

#common library used for graph plotting in R
library(ggplot2)

#libraries specifically needed for raincloud-plot
library(cowplot)
library(dplyr)
library(readr)

#specify script for running raincloud-plot ('source' function read R code from a file)
#this 'R_rainclouds.R' script was downloaded from the official RainCloudPlots repository:  https://github.com/RainCloudPlots/RainCloudPlots/tree/master/tutorial_R

source("R_rainclouds.R")
```


Let's define a theme to make pretty plots
```{r}

raincloud_theme <- theme(
  #axis.title.x =  element_blank(),
  #axis.title.y =  element_blank(), 
  axis.text = element_text(size = 14, face="bold", color="black"),
  legend.title=element_text(size=16),
  legend.text=element_text(size=16),
  legend.position = "right",
  panel.border = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  panel.grid.major = element_blank(),
  axis.line.x = element_line(colour = 'black', size=0.5, linetype='solid'),
  axis.line.y = element_line(colour = 'black', size=0.5, linetype='solid'))


```


Summarising the data: in this plot we are interested in the average stimulus rating for each participant, separately for category (food vs curiosity) and decision (accept vs reject)

```{r}

data_by_subj <- aggregate(raw_rate~participant+category+choice, data=df, FUN=mean)

data_by_subj["Decision"] <-NA
data_by_subj$Decision[data_by_subj$choice==0] <- "Reject"
data_by_subj$Decision[data_by_subj$choice==1] <- "Accept"

```


let's also calculate some summary statistics for each condition (i.e. 1.food-accept, 2.food-reject, 3.curiosity-accept, 4.curiosity-reject), including mean, S.D., & S.E. of rating. They will be used in plotting geom_pointrange

```{r}

gbq1 <- aggregate(raw_rate~category+choice, data=df, FUN=mean)
names(gbq1)[3]<- "mean_rate"
gbq2 <-  aggregate(raw_rate~category+choice, data=df, FUN=sd)  
names(gbq2)[3]<- "sd_rate"
gbq<- merge(gbq1, gbq2, by = c("category", "choice"))

data.frame(gbq)
names(gbq)=c("category","choice","mean_rate","sd_rate")

gbq["se_rate"] <- gbq$sd_rate/sqrt(length(unique(df$participant)))

gbq["Decision"] <-NA
gbq$Decision[gbq$choice==0] <- "Reject"
gbq$Decision[gbq$choice==1] <- "Accept"

gbq$ymin = with(gbq, mean_rate - se_rate)
gbq$ymax = with(gbq, mean_rate + se_rate)
```

Ready to plot?!
```{r}

g <- ggplot() + 
  geom_flat_violin(data = data_by_subj, aes(y = raw_rate, x = category, fill=Decision), position = position_nudge(x = .2, y = 0), alpha = .7, colour=FALSE, show.legend = FALSE) +
  geom_point(data=data_by_subj, aes(y = raw_rate, x = category, color = Decision), position = position_jitter(width=.1),size = 2, alpha = 1, show.legend = FALSE, shape=1) +
  geom_pointrange (data=gbq, aes(y = mean_rate, x = category, ymin= ymin, ymax= ymax, color=Decision), shape=16, size=1, position = position_dodge(width=0.1)) + 
  scale_x_discrete(labels=c("Curiosity","Food")) +
  ylim(0,7) +
  
  #guides(fill=guide_legend(title="Choices")) +
  raincloud_theme +
  labs(x=NULL, y="Stimulus Rating")

g

```

If you want to save the plot as an image
```{r}

ggsave("rainplot_trial.jpg", width = 20, height = 20, units = "cm", dpi=300)

```
